NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-11-30 Manuscript/manuscriptPGPE.tex
Update manuscript

@@

Derivation of different lipid headgroup conformational ensembles in liquid state has been
%discussed previously \cite{??}, but these studies have been
inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.
Our experimental order parameters, including the signs, from different lipid headgroups
combined with the literature data are shown in Fig. \ref{HGorderParameters}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG. 
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
While the headgroup order parameters are similar for PC and PE lipids,
PG and PS lipids exhibit distinct values in comparison with other lipids.
In PS headgroup, the $\alpha$-carbon order parameter exhibits significant forking and
the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup, the $\alpha$-carbon order parameter is similar to PE and PG,
while the positive value of $\beta$-carbon is distinct from all the other lipids.
Notably, this difference was not observed in previous $^2$H NMR experiments,
because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and the order parameter
signs were not measured~\cite{wohlgemuth80,gally81,borle85}.

@@

As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
none of the MD simulation force fields correctly captures all the
headgroup and glycerol backbone order parameters of PE and PG lipids
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
that would enable a straightforward interpretation of conformational ensembles.
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
captures the essential differences between
PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
which is too negative when compared with other lipids or experiments \cite{botan15}.
% As already pointed out in our previous study \cite{antila19},

The conformational ensembles characterized by the distributions of heavy atom dihedral angles
show major differences between lipids only for O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds
near the end of the headgroup (Fig. \ref{structures}).
These probably explain the slight differences in the
angle between headgroup dipole and membrane normal, which decreases 
in the order PG $>$ PE  $>$ PC  $>$ PS, although difference between PC and PE may an artefact as
the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field.
Dihedrals of both O-P bonds can freely rotate in all lipids as all possible angles
are observed in the distributions. For PS, however, some angles are quite unlikely,
possible explaining the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.
The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

@@

\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/PCPEPGPSstructures.png}
  \includegraphics[width=8.0cm]{../Figs/CHARMMfromLIPIDS.eps}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals1.png}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals2.png}
  \includegraphics[width=8.0cm]{../Figs/PNangleNORM.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALS.eps}
  \caption{\label{structures}
    Overlayed snapshots and dihedral angle distributions from CHARMM36 simulations of different lipids
    which give the best agreement with experiments.
  }
  \todo{The differences observed in dihedral distributions are not visible in the snapshot figures.} \\
  \todo{More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}

>>>>

Derivation of different lipid headgroup conformational ensembles in liquid state has been
inconclusive due to lack of suitable experimental data and tools to interpret the conformational ensembles.

@@

To experimentally characterize the differences in lipid headgroup conformational ensembles
in liquid bilayer phase, we measured the C-H bond order parameters magnitudes and signs
from POPG and POPE bilayers, and compared those to our previously published results
for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters}.
%Our experimental order parameters, including the signs, from different lipid headgroups
%combined with the literature data are shown in Fig. \ref{HGorderParameters}.
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.
The most distinct order parameters are observed for PS and PG headgroups.
In PS, the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup,
%the $\alpha$-carbon order parameter is similar to PE and PG, while
the $\beta$-carbon order parameter has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.

@@

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different lipids,
see Refs. \cite{botan15} and \cite{antila19}, and Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}.
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).

show major differences between lipids only for O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$ bonds
near the end of the headgroup (Fig. \ref{structures}).
These probably explain the slight differences in the
angle between headgroup dipole and membrane normal, which decreases 
in the order PG $>$ PE  $>$ PC  $>$ PS, although difference between PC and PE may an artefact as
the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field.
As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
none of the MD simulation force fields correctly captures all the
headgroup and glycerol backbone order parameters of PE and PG lipids
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
that would enable a straightforward interpretation of conformational ensembles.
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
captures the essential differences between
PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
which is too negative when compared with other lipids or experiments \cite{botan15}.
% As already pointed out in our previous study \cite{antila19},

Dihedrals of both O-P bonds can freely rotate in all lipids as all possible angles
are observed in the distributions. For PS, however, some angles are quite unlikely,
possible explaining the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.
The eclipsed (0 or 360 degrees) are not present in any other dihedrals, and anti-eclipsed (120 or 240 degrees)
are not present in C$_\alpha$-C$_\beta$ and g$_1$-g$_2$ bonds.

@@

\begin{figure*}[!h]
  \centering
  \includegraphics[width=18.0cm]{../Figs/PCPEPGPSstructures2.png}
  \includegraphics[width=8.0cm]{../Figs/CHARMMfromLIPIDS.eps}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals1.png}
%  \includegraphics[width=8.0cm]{../Figs/PCPEPGPSdihedrals2.png}
  \includegraphics[width=8.0cm]{../Figs/PNangleNORM.eps}
  \includegraphics[width=8.0cm]{../Figs/DIHEDRALS.eps}
  \caption{\label{structures}
    Overlayed snapshots and dihedral angle distributions from CHARMM36 simulations of different lipids
    which give the best agreement with experiments.
  }
  \todo{More detailed discussion of this figure is in https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/9} \\
\end{figure*}
